A Parallel General Purpose Multi-Objective Optimization Framework, with Application To Beam Dynamics

### 2018-03-16 paper/theor-experiment.tex
fixed table

\begin{table}[h!]
\begin{center}
  \caption{Convergence of benchmark problem with errors relative to
    hypervolume of sampled reference solution.}
  \label{tbl:bench_rms_error}
  \begin{tabular}{lcc}
    \hline\noalign{\smallskip}
    tot.\ function  & hyper volume & relative error\\
    evaluations    & & \\
    \noalign{\smallskip}\hline\noalign{\smallskip}
    100  &  0.859753 & $3.076 \times 10^{-1}$ \\
    \noalign{\smallskip}\hline\noalign{\smallskip}
    200  &  0.784943 & $1.938 \times 10^{-1}$ \\
    500  &  0.685183 & $4.210 \times 10^{-2}$ \\
    900  &  0.661898 & $6.689 \times 10^{-3}$ \\
    1100 &  0.657615 & $1.749 \times 10^{-4}$ \\
    \noalign{\smallskip}\hline
  \end{tabular}
\end{center}
\end{table}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.8\linewidth]{Report/dt_scan}
  \caption{Envelope-tracker with different number of slices and simulation
    time steps.}
  \label{fig:et-dt}
\end{figure}

@@

\begin{figure}[h!]
  \centering
    \includegraphics[width=0.8\linewidth]{Report/front_plot}
  \caption{Pareto front for the $1000$th generation with $40$ individuals
    using 400 slices (interesting region magnified), a simulation timestep
    of 1.5 ps.
    The individual 3 was selected for further investigations.}
  \label{fig:pareto_front_21}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.8\linewidth]{Report/iff_plot_emrms}
  \caption{Beamsize and emittance of individual $3$.}
  \label{fig:rmsemit}
\end{figure}

@@

For validation purposes we compared the results of the envelope-tracker using
  the analytical space charge model with the \textsc{OPAL} 3D macro particle
  tracker.
The benchmark was run on the first 12.5 meters of the \textsc{SwissFEL} 250
  MeV injector.
The results for both rms beamsize and emittance are shown in
  Figure~\ref{fig:et-vs-tt}.
A good agreement between the two codes can be observed.
The difference of the larger emittance along the solenoids in case of 3D
  tracker that is not seen by the envelope-tracker is due to the different
  definition of the particle momenta (canonical vs.~mechanical).
Both trackers agree within acceptable limits \cite{chao:99}.
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{Report/iff_plot_emrms_ettt}
  \caption{Comparison 3D-tracker versus envelope-tracker in case for rms$_{x}$
    and $\varepsilon_{x}$.}
  \label{fig:et-vs-tt}
\end{figure}

@@

%Currently we run the simulations on a low resolution space charge grid ($8
  %\times 8 \times 8$) as a proof of concept.
%In a next step a simple Python script would be applied to iteratively rerun
  %the optimizer with increasing space-charge resolution when the limits of the
  %design variables are narrow enough (iteratively diminishing search space).

\subsection{AWA Photoinjector Optimization}
Next we apply the optimization framework to the high charge beam line at the AWA, 
which consists of an rf photocathode gun, two solenoids, six linear accelerating cavities, 
followed by four quadrupoles. This beam line has served as the injector for Two Beam 
Acceleration (TBA) experiments in the past [XX], and will be used in the future for 
similar experiments. One of the main constraints on prior experimental 
results was the beam size and emittance as the beam passed through small aperture 
wakefield structures. In an attempt to maximize charge transmission in future experiments
we optimize several parameters in the gun and quadrupole leading into the 
TBA section of the beam line.
\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{Report/awa-drawing}
	\caption{High charge beam line layout at the AWA.}
	\label{awa-pic}
\end{figure}

Compared to the Ferrario experiment in Section \ref{ferrario}, 
we increase the number of design variables and objectives by a factor of two.
The problem is also less predictable, due to the large number of design variables
and nonlinear effects such as space charge, there is not a definite idea
before hand where the Pareto front will fall. This work is also meaningful in that
it will guide future operations at the AWA.

We define the AWA optimization problem using objectives, and parameters given in 
equations (\ref{eq:awa:p1}) to (\ref{eq:awa:lastdvar}).

\begin{align}
\text{min}  \quad & \text{rms}_{x, s1}, \quad \text{rms}_{x, s2} \label{eq:awa:p1}\\
& \text{rms}_{s, s1}, \quad \text{rms}_{s, s2} \label{eq:awa:p2}\\
& \varepsilon_{x,s1}, \quad \varepsilon_{x,s2} \label{eq:awa:p3} \\
\text{subject to} \quad & q = 40 \left[\text{nC}\right] \label{eq:awa:firstconstr}\\
\quad & \text{Volt}_{\text{Gun}} = 64\left[\text{MV/m}\right] \label{eq:awa:lastconstr}\\
\quad & \text{Volt}_{\text{Linac}} = 24-25\left[\text{MV/m}\right] \\
\quad & \sigma_{L} \leq \sigma_x = \sigma_y \leq \sigma_{U} \label{eq:awa:firstdvar}\\
\quad & \text{KS}_{L} \leq \text{KS}_{\text{RF}} \leq \text{KS}_{U} \label{eq:awa:seconddvar}\\
\quad & \text{LAG}_{L} \leq \text{LAG}_{\text{RF}} \leq \text{LAG}_{U} \\
\quad & \Delta z_{L\text{KS}} \leq \Delta z_{\text{KS}} \leq \Delta z_{U\text{KS}} \label{eq:awa:lastdvar}
\end{align}

The first four objectives minimize the transverse ($rms_x$) and longitudinal ($rms_s$) 
beam sizes at to locations of interest in the beam line, [$s_1$,$s_2$]. 
These are the longitudinal positions of 
the entrance of the kicker and exit of the kicker. The beam size needs to be minimized
here to prevent scraping and ensure transmission through the wakefield structures 
following the kicker. The last two objectives (\ref{eq:awa:p3}) minimize the 
transverse emittance, which favors less divergence in the beam.

@@

@@

 \vspace{0.2cm}
 {\footnotesize \begin{verbatim}
rmss1:  OBJECTIVE,EXPR="fabs(statVariableAt('rms_s',16.45))";
rmsx1:  OBJECTIVE,EXPR="fabs(statVariableAt('rms_x',16.45))";
emitx1: OBJECTIVE,EXPR="fabs(statVariableAt('emit_x',16.45))";

rmss2:  OBJECTIVE,EXPR="fabs(statVariableAt('rms_s',18.45))";
rmsx2:  OBJECTIVE,EXPR="fabs(statVariableAt('rms_x',18.45))";
emitx2: OBJECTIVE,EXPR="fabs(statVariableAt('emit_x',18.45))";

de1:    OBJECTIVE,EXPR="fabs(statVariableAt('dE',16.45))";

All simulations for this experiment were carried out on Bebop a
high performance computing (HPC)
cluster provided by the Laboratory Computing Resource Center (LCRC)
at Argonne National Laboratory (ANL). We used Intel Knights Landing 
(KNL) processors at 1.3 GHz with 128 GB of memory per node, 
and 64 cores per node. There are 352 compute nodes available on 
Bebop, with a total of 22528 cores. 
This allows for very large optimization jobs, like the AWA case.

OPAL-T was used as the forward solver, and a study on time step and 
number of particles was done to reduce the time to simulation while 
maintaining the physics of interest. All jobs were run and compared 
on 8 cores to take advantage of the KNL, by running 8 jobs per node.
After comparing several options we choose ten thousand particles 
in combination with adjusting the time step near sensitive elements 
(i.e. quadurpoles, dipole). 
The resulting simulations are low fidelity, but closely approximate 
the mid fidelity simulations for metrics of interest, as shown in 
Fig. \ref{tstep}.

\begin{table}[h!]
	\begin{center}
		\caption{Low and mid fidelity simulation parameters.}
		\label{fidelity}
		\begin{tabular}{l ccc}
			\hline\noalign{\smallskip}
			 & Number of particles & Time step (seconds) & Time to simulation (minutes)\\
			\noalign{\smallskip}\hline\noalign{\smallskip}
			low fidelity  			&  10,000   & $5 \times10^{-11}$  &  2.6 \\
			mid fidelity 			&  100,000  & $1 \times 10^{-12}$ &  \\
			adjusted low fidelity  	&  10,000   & $2 \times 10^{-12}$, $1 \times 10^{-11}$ & 1.5 \\
			\noalign{\smallskip}\hline
		\end{tabular}
	\end{center}
\end{table}

@@

>>>>

\begin{table}%[h!]
\begin{center}
  \caption{Convergence of benchmark problem with errors relative to
    hypervolume of sampled reference solution.}
  \label{tbl:bench_rms_error}
  \begin{tabular}{lcc}
    \hline\noalign{\smallskip}
    tot.\ function  & hyper volume & relative error\\
    evaluations    & & \\
    \noalign{\smallskip}\hline\noalign{\smallskip}
    100  &  0.859753 & $3.076 \times 10^{-1}$ \\
    \noalign{\smallskip}\hline\noalign{\smallskip}
    200  &  0.784943 & $1.938 \times 10^{-1}$ \\
    500  &  0.685183 & $4.210 \times 10^{-2}$ \\
    900  &  0.661898 & $6.689 \times 10^{-3}$ \\
    1100 &  0.657615 & $1.749 \times 10^{-4}$ \\
    \noalign{\smallskip}\hline
  \end{tabular}
\end{center}
\end{table}

@@

\begin{figure}%[h!]
  \centering
  \includegraphics[width=0.8\linewidth]{Report/dt_scan}
  \caption{Envelope-tracker with different number of slices and simulation
    time steps.}
  \label{fig:et-dt}
\end{figure}

@@

\begin{figure}%[h!]
  \centering
    \includegraphics[width=0.8\linewidth]{Report/front_plot}
  \caption{Pareto front for the $1000$th generation with $40$ individuals
    using 400 slices (interesting region magnified), a simulation timestep
    of 1.5 ps.
    The individual 3 was selected for further investigations.}
  \label{fig:pareto_front_21}
\end{figure}

@@

\begin{figure}%[h!]
  \centering
  \includegraphics[width=0.8\linewidth]{Report/iff_plot_emrms}
  \caption{Beamsize and emittance of individual $3$.}
  \label{fig:rmsemit}
\end{figure}

@@

For validation purposes we compared the results of the envelope-tracker using
  the analytical space charge model with the \textsc{OPAL} 3D macro particle
  tracker.
The benchmark was run on the first 12.5 meters of the \textsc{SwissFEL} 250
  MeV injector.
The results for both rms beamsize and emittance are shown in
  Figure~\ref{fig:et-vs-tt}.
A good agreement between the two codes can be observed.
The difference of the larger emittance along the solenoids in case of 3D
  tracker that is not seen by the envelope-tracker is due to the different
  definition of the particle momenta (canonical vs.~mechanical).
Both trackers agree within acceptable limits \cite{chao:99}.

\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{Report/iff_plot_emrms_ettt}
  \caption{Comparison 3D-tracker versus envelope-tracker in case for rms$_{x}$
    and $\varepsilon_{x}$.}
  \label{fig:et-vs-tt}
\end{figure}

@@

%Currently we run the simulations on a low resolution space charge grid ($8
  %\times 8 \times 8$) as a proof of concept.
%In a next step a simple Python script would be applied to iteratively rerun
  %the optimizer with increasing space-charge resolution when the limits of the
  %design variables are narrow enough (iteratively diminishing search space). 
%\vspace{-7em}
\subsection{AWA Photoinjector Optimization}
Next we apply the optimization framework to the high charge beam line at the AWA, 
which consists of an rf photocathode gun, two solenoids, six linear accelerating cavities, 
followed by four quadrupoles and a stripline kicker, as shown in Fig. \ref{awa-pic}. 
Portions of this beam line served as the injector for Two Beam 
Acceleration (TBA) experiments in the past [X1], and will be used for 
similar experiments in the future [X2]. Main constraints on prior experimental 
results were the beam size and emittance as the beam passed through small aperture 
wakefield structures located after the quadrupoles. 
In an attempt to maximize charge transmission in upcoming experiments
we optimize several parameters in the gun and quadrupole leading into the 
TBA section of the beam line.
%\vspace{-1em}

\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{Report/awa-drawing}
	\caption{High charge beam line layout at the AWA.}
	\label{awa-pic}
\end{figure}

\vspace{-1em}
Compared to the Ferrario experiment in Section \ref{ferrario}, 
we increase the number of design variables and objectives by a factor of two.
The problem is also less predictable, due to the increased number of design variables
and nonlinear effects such as space charge, there is not a definite idea
before hand where the Pareto front will fall. This work is also meaningful in that
it will guide future operations at the AWA.

\begin{align}
\text{min}  \quad & \text{rms}_{x, s_1}, \quad \text{rms}_{x, s_2} \label{eq:awa:p1}\\
& \text{rms}_{s, s_1}, \quad \text{rms}_{s, s_2} \label{eq:awa:p2}\\
& \varepsilon_{x,s_1}, \quad \varepsilon_{x,s_2} \label{eq:awa:p3} \\
& dE \label{eq:awa:p4} \\
\text{subject to} \quad & q = 40 \left[\text{nC}\right] \label{eq:awa:firstconstr}\\
\quad & \text{Volt}_{\text{Gun}} = 64\left[\text{MV/m}\right] \label{eq:awa:lastconstr}\\
\quad & \text{Volt}_{\text{Linac}} = 24-25\left[\text{MV/m}\right] \\
\quad & \sigma_x = \sigma_y = 9 \left[\text{mm}\right] \label{eq:awa:firstdvar}\\
\quad & \phi_{\text{linac}} =0 \label{eq:awa:lastdvar}
\end{align}

The first four objectives minimize the transverse ($rms_x$) and longitudinal ($rms_s$) 
beam sizes at two locations of interest in the beam line ($s_1$,$s_2$). 
These are the longitudinal positions at the entrance of the kicker and 
$1[\text{m}]$ after the exit of the kicker. The beam size needs to be minimized
here to prevent scraping and ensure transmission through the wakefield structures 
following the kicker. The next two objectives (\ref{eq:awa:p4}) minimize the 
transverse emittance, which favors less divergence in the beam and reduces 
emittance growth during transport. The last objective (\ref{eq:awa:p4}), 
minimizes energy spread in the bunch at the entrance of the kicker, $s_1$.  
This helps reduce emittance and beam size growth in the kicker.  
Equations (\ref{eq:awa:firstconstr}) to
(\ref{eq:awa:lastdvar}) define the charge, gun voltage, linac voltages, 
laser radius and linac cavity phases. These are parameters in the simulation 
that must be defined, but do not vary during the optimization.

%\vspace{-1em}
As introduced in Section \ref{ferrario}, we define the AWA optimization problem 
objectives, and design variables in the OPAL input file as shown in
the following code: %equations (\ref{eq:awa:p1}) to (\ref{eq:awa:lastdvar}).

@@

	// Objectives
	rmss1:  OBJECTIVE,EXPR="fabs(statVariableAt('rms_s',16.45))";
	rmsx1:  OBJECTIVE,EXPR="fabs(statVariableAt('rms_x',16.45))";
	emitx1: OBJECTIVE,EXPR="fabs(statVariableAt('emit_x',16.45))";
	de1:    OBJECTIVE,EXPR="fabs(statVariableAt('dE',16.45))";	
	rmss2:  OBJECTIVE,EXPR="fabs(statVariableAt('rms_s',18.45))";
	rmsx2:  OBJECTIVE,EXPR="fabs(statVariableAt('rms_x',18.45))";
	emitx2: OBJECTIVE,EXPR="fabs(statVariableAt('emit_x',18.45))";

@@

Design variables include the currents in two gun solenoids (IBF and IM), 
the gun phase (GPHASE), the laser longitudinal full width half max (FWHM), 
and four quadrupole strengths (KQ1-KQ2). The objectives include
beam size, emittance, and energy spread as
defined in Eqs. (\ref{eq:awa:p1}) to (\ref{eq:awa:p4}). 
The s location at the entrance of the kicker is $s_1=16.45[\text{m}]$, 
and the s location of interest after the kicker is $s_1=18.45[\text{m}]$.

All simulations for this experiment were carried out on Bebop a
high performance computing (HPC)
cluster provided by the Laboratory Computing Resource Center (LCRC)
at Argonne National Laboratory (ANL). We used Intel Knights Landing 
(KNL) processors at 1.3 GHz with 128 GB of memory 
and 64 cores per node. There are 352 compute nodes available on 
Bebop, with a total of 22,528 cores. 
This allows for very large optimization jobs, like the AWA case.

OPAL was used as the forward solver, and a study on time step and 
number of particles was done to reduce the time to simulation while 
maintaining the physics of interest. All jobs were run and compared 
on 8 cores each, which allowed 8 jobs per node on the KNL.
The grid size was $16 \times 16 \times 32$, 
and parallelized in the x and y directions.
After comparing several options we choose ten thousand particles 
in combination with adjusting the time step near sensitive elements 
(i.e. quadurpoles, kicker). 
The resulting simulations are low fidelity, but closely approximate 
the mid fidelity simulations for metrics of interest, as shown in 
Fig. \ref{tstep}.

@@

\begin{table}%[h!]
	\begin{center}
		\caption{Low and mid fidelity simulation parameters.}
		\label{fidelity}

		\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} C c D }
			\hline\noalign{\smallskip}
			& Number of particles & dT (seconds) & Time to simulation (minutes)\\
			\noalign{\smallskip}\hline\noalign{\smallskip}
			low fidelity  			&  10,000   & $5 \times10^{-11}$  &  2.6 \\
			adjusted low fidelity  	&  10,000   & $2 \times 10^{-12}$, $1 \times 10^{-11}$ & 1.6 \\
			mid fidelity 			&  100,000  & $1 \times 10^{-12}$ &  18\\
			\noalign{\smallskip}\hline
		\end{tabular*}
	\end{center}
\end{table}
